\Subsection{conc-ffi}{Concurrency and the Foreign Function Interface}

Haskell has a \emph{foreign function interface} (FFI) that allows
Haskell code to call, and be called by, foreign language code
(primarily C) \cite{haskell2010}.  Foreign languages also have their
own threading models --- in C there is POSIX or Win32 threads, for
example --- so we need to specify how Concurrent Haskell interacts
with the threading models of foreign code.

The details of the design can be found in \citet{conc-ffi}, in the
following sections we summarise the behaviour the Haskell programmer
can expect.

All of the following assumes that GHC's @-threaded@ option is in use.
Without @-threaded@, the Haskell process uses a single OS thread only,
and multi-threaded foreign calls are not supported.

\Subsubsection{conc-ffi-outcall}{Threads and foreign out-calls}

An out-call is a call made from Haskell to a foreign language.  At the
present time the FFI supports only calls to C, so that's all we
describe here.  In the following we refer to threads in C (i.e. POSIX
or Win32 threads) as ``OS threads'' to distinguish them from Haskell
threads.

As an example, consider making the POSIX C function @read()@ callable
from Haskell:

\begin{haskell}
foreign import ccall "read"
   c_read :: CInt      -- file descriptor
          -> Ptr Word8 -- buffer for data
          -> CSize     -- size of buffer
          -> CSSize    -- bytes read, or -1 on error
\end{haskell}

\noindent This declares a Haskell function @c_read@ that can be used
to call the C function @read()@.  Full details on the syntax of
@foreign@ declarations and the relationship between C and Haskell
types can be found in the Haskell report \cite{haskell2010}.

Just as Haskell threads run concurrently with each other, when a
Haskell thread makes a foreign call, that foreign call runs
concurrently with the other Haskell threads, and indeed with any other
active foreign calls.  Clearly the only way that two C calls can be
running concurrently is if they are running in two separate OS
threads, so that is exactly what happens: if several Haskell threads
call @c_read@ and they all block waiting for data to be read, there
will be one OS thread per call blocked in @read()@.

This has to work despite the fact that Haskell threads are not
normally mapped one-to-one with OS threads; as we mentioned earlier
(\secref{forking}), in GHC, Haskell threads are lightweight and
managed in user-space by the runtime system.  So to handle concurrent
foreign calls, the runtime system has to create more OS threads, and
in fact it does this on demand.  When a Haskell thread makes a foreign
call, another OS thread is created (if necessary), and the
responsibility for running the remaining Haskell threads is handed
over to the new OS thread, meanwhile the current OS thread makes the
foreign call.

The implication of this design is that a foreign call may be executed
in \emph{any} OS thread, and subsequent calls may even be executed in
different OS threads. In most cases this isn't important, but
sometimes it is: some foreign code must be called by a \emph{particular}
OS thread.  There are two instances of this requirement:

\begin{itemize}
\item Libraries that only allow one OS thread to use their API.  GUI
  libraries often fall into this category: not only must the library
  be called by only one OS thread, it must often be one
  \emph{particular} thread (e.g. the main thread).  The Win32 GUI APIs
  are an example of this.

\item APIs that use internal thread-local state.  The best-known
  example of this is OpenGL, which supports multi-threaded use, but
  stores state between API calls in thread-local storage.  Hence,
  subsequent calls must be made in the same OS thread, otherwise the
  later call will see the wrong state.
\end{itemize}

For this reason, the concept of \emph{bound threads} was introduced.
A bound thread is a Haskell thread/OS thread pair, such that foreign
calls made by the Haskell thread always take place in the associated
OS thread.  A bound thread is created by @forkOS@:

\begin{haskell}
forkOS :: IO () -> IO ThreadId
\end{haskell}

\noindent Care should be taken when calling @forkOS@: it creates a
complete new OS thread, so it can be quite expensive.

\Subsubsection{conc-ffi-incall}{Threads and foreign in-calls}

In-calls are calls to Haskell functions that have been exposed to
foreign code using @foreign export@.  For example, if we have a
function @f@ of type @Int -> IO Int@, we could expose it like this:

\begin{haskell}
foreign export ccall "f" f :: Int -> IO Int
\end{haskell}

\noindent This would create a C function with the following signature:

\begin{haskell}
HsInt f(HsInt);
\end{haskell}

\noindent here @HsInt@ is the C type corresponding to Haskell's @Int@
type.

In a multi-threaded program, it is entirely possible that @f@ might be
called by multiple OS threads concurrently.  The GHC runtime system
supports this (at least with @-threaded@), with the following
behaviour: each call becomes a new \emph{bound thread}.  That is, a
new Haskell thread is created for each call, and the Haskell thread is
bound to the OS thread that made the call.  Hence, any further
out-calls made by the Haskell thread will take place in the same OS
thread that made the original in-call.  This turns out to be important
for dealing with GUI callbacks: the GUI wants to run in the main OS
thread only, so when it makes a callback into Haskell, we need to
ensure that GUI calls made by the callback happen in the same OS
thread that invoked the callback.

\Subsubsection{conc-ffi-further}{Further reading}

\begin{itemize}
\item The full specification of the Foreign Function Interface (FFI)
  can be found in the Haskell 2010 report \cite{haskell2010};
\item GHC's extensions to the FFI can be found in the GHC User's
  Guide\footnote{\url{http://www.haskell.org/ghc/docs/latest/html/users_guide/}};
\item Functions for dealing with bound threads can be found in the
  documentation for the @Control.Concurrent@ module.
\end{itemize}

